Assessing primitives performance\\ on multi-stage execution

### 2018-05-16 paper.tex
add a bunch of comments

\cba{Cette partie je faisais la queue dans la gare en mm temps, a relire}
\sd{je ne suis pas sur que je garderais la specification originale car cela n'apporte pas grand-chose.}
 The primitive specification is a bit tedious to explain, since as we were trying to improve its performance, we had to change the specifications. In this section we describe first the string implementation, then the original primitive specification and why they were problematic and lastly the new specifications.

@@

\paragraph{Original specification.}  The original primitive took three operands. The two first operands were the two ByteStrings to compare and the third operand was the order table, a ByteArray encoding the order of characters. The default order is the ASCII order, hence each entry corresponds to its index in the byte array (The ByteArray has a size of 256, and accessing the element i, with 0 $\leq$ i $\leq$ 255, answered i). However, in some case, the program may do case insensitive ByteStrings comparison. In this case, the order ByteArray is different, since some entries answer the same value (for example, the entry for the 41 and 61, respectively characters A and a, both answer the same value). To compare two ByteStrings, the primitive conceptually iterates byte by byte over the two objects until it reaches the end of one them, comparing for each byte that the value present in the order ByteArray at the index is greater, equal or smaller than the value computed from the other ByteString. If a difference is met, the value returns based on the result, if not, at the end of the iteration the primitive compares the ByteString sizes and returns. Based on the three parameters, the primitive answered 3, 2, or 1 depending if the first ByteString was greater, equal or smaller than the second one.

\paragraph{Specification issues.} As we were trying to optimize string comparison but specifically string equality, we encountered two main issues:
\begin{itemize}
	\item \emph{Returned value:} Other programming languages implement the primitive to return a negative integer, 0 or a positive integer instead of 1, 2 or 3. It looks like a minor difference, but it is important for performance. With a negative, 0 or positive value, the primitive can just answer the difference between two different characters or the difference in size if all first characters are equal. With 1, 2, or 3, the primitive has to end with multiple branches to know what number to return. In native code, a subtraction is on common architecture quicker to perform than one or multiple branches. In addition, the primitive code is easier to write with the subtraction when writing directly the assembly instructions and we did exactly that when writing the primitive on top of the JIT's back-end.
	\item \emph{Order:} Having the order as the third operands is convenient in some cases (case insensitive comparison), but is just overhead in the common case (checking for string equality).
\end{itemize}

\paragraph{Final specification.} The final specification we productized last month takes two or three parameters. The third parameter, the order, is optional. If the third parameter is absent, the VM assumes the byte ordering (ASCII order) is the order to compare against. Therefore, the VM does not need to go through the indirection order array in the common case. The primitive answers a negative integer, zero, or a positive integer as the first parameter is less than, equal to, or greater than the second parameter.

@@

 This section first describes the different ways of implementing a primitive. We then explain how each of these versions is executed by the VM and the pros and cons of each and every.

 \subsection{Different implementations and executions}
We distinguish several implementations: the first one

@@

\paragraph{Existing: Smart Syntax Plugin.}

@@

\paragraph{Existing: Slang Plugin.}
The Slang plugin is a recent complete rewrite of the Smart Syntax plugin. The main difference is that the Smalltalk version is used only as fall-back code, while at Slang-to-C compilation time the VM uses code written in Slang to generate the C code. One of the side-effect is that the plugin's code is now part of the VM code base and not the client code bases, which is convenient for maintenance since the Cog VM has multiple client languages. However, the implementor now needs to maintain and evolve two version of the primitive, the fall-back code in Smalltalk and the code in Slang. \sd{do we need the next sentence?}This primitive follows the old specification, the full code is available in Appendix~\ref{PluginAppendix}.

@@

\paragraph{Pure Slang with JIT support.}\sd{Why not calling it Slang+RTL}

@@

 In this section we compare the five different implementations according to the following criteria:
 \begin{itemize}
 	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes). \sd{why not ST, Plugin, VM?}
	\item \emph{Implementation language familiar to developers:} Smalltalk application developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code. Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.  We use All if all communities are familiar with the language, VM if all VM developers are and VM-JIT if only a subset of VM developers are.
	\item \emph{Readability:} To evaluate the approach readability, we use the number of lines of code even if this is arguable but used regularly in other papers \cite{xxx}. A better metric would be to use the cyclomatic complexity, however since the  Cog' s RTL generates branches at runtime, it is difficult to use it.
	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to be executed if the primitive fails. This code is maintained by the client language implementors, and not by the VM implementors. Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation\sd{why?}. However, as a VM implementor, we need to maintain what generates the C code of the VM, hence, in the case of Smart Syntax, we need to maintain a different method in each client language. We also count all the Slang implementations as well as the ones on Cog's RTL.
	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
	\item \emph{Optimization potential on large string comparison:}
	C cross-file inlining is not performed by default for internal plugins. External plugins don't have access to the whole VM API and have to go through an interface to do so, leading to an overhead.
	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code. The generated code can therefore differ according the compilers and the optimizations they performed. On the contrary, the JIT will always generate the same native code.
\end{itemize}

>>>>

\cba{Cette partie je faisais la queue dans la gare en mm temps, a relire}
\sd{je ne suis pas sur que je garderais la specification originale car cela n'apporte pas grand-chose.}
 The primitive specification is a bit tedious to explain, since as we were trying to improve its performance, we had to change the specifications. \sk{rephrase bc subsections deleted} In this section we describe first the string implementation, then the original primitive specification and why they were problematic and lastly the new specifications.

@@

\paragraph{Original specification.}  \sk{delete}The original primitive took three operands. The two first operands were the two ByteStrings to compare and the third operand was the order table, a ByteArray encoding the order of characters. The default order is the ASCII order, hence each entry corresponds to its index in the byte array (The ByteArray has a size of 256, and accessing the element i, with 0 $\leq$ i $\leq$ 255, answered i). However, in some case, the program may do case insensitive ByteStrings comparison. In this case, the order ByteArray is different, since some entries answer the same value (for example, the entry for the 41 and 61, respectively characters A and a, both answer the same value). To compare two ByteStrings, the primitive conceptually iterates byte by byte over the two objects until it reaches the end of one them, comparing for each byte that the value present in the order ByteArray at the index is greater, equal or smaller than the value computed from the other ByteString. If a difference is met, the value returns based on the result, if not, at the end of the iteration the primitive compares the ByteString sizes and returns. Based on the three parameters, the primitive answered 3, 2, or 1 depending if the first ByteString was greater, equal or smaller than the second one.

\paragraph{Specification issues.} \sk{delete} As we were trying to optimize string comparison but specifically string equality, we encountered two main issues:
\begin{itemize}
	\item \emph{Returned value:} Other programming languages implement the primitive to return a negative integer, 0 or a positive integer instead of 1, 2 or 3. It looks like a minor difference, but it is important for performance. With a negative, 0 or positive value, the primitive can just answer the difference between two different characters or the difference in size if all first characters are equal. With 1, 2, or 3, the primitive has to end with multiple branches to know what number to return. In native code, a subtraction is on common architecture quicker to perform than one or multiple branches. In addition, the primitive code is easier to write with the subtraction when writing directly the assembly instructions and we did exactly that when writing the primitive on top of the JIT's back-end.
	\item \emph{Order:} Having the order as the third operands is convenient in some cases (case insensitive comparison), but is just overhead in the common case (checking for string equality).
\end{itemize}

\paragraph{Final specification.} \sk{rename and possibly rephrase}The final specification we productized last month takes two or three parameters. The third parameter, the order, is optional. If the third parameter is absent, the VM assumes the byte ordering (ASCII order) is the order to compare against. Therefore, the VM does not need to go through the indirection order array in the common case. The primitive answers a negative integer, zero, or a positive integer as the first parameter is less than, equal to, or greater than the second parameter.

@@

\sk{So, do we agree that: 1- we bench all implem with new spec (i am rewriting the plugins right now) 2- we drop the sections "original spec", "spec issues", rename "final spec" and rephrase the implementations details on the sections below?}
 This section first describes the different ways of implementing a primitive. We then explain how each of these versions is executed by the VM and the pros and cons of each and every.

 \subsection{Different implementations and executions}
We distinguish several implementations: the first one \sk{delete "the first one"? sounds odd}

@@

\paragraph{Existing: Smart Syntax.}

@@

\paragraph{Existing: Plugin.}
The Slang plugin is a recent complete rewrite of the Smart Syntax plugin. The main difference is that the Smalltalk version is used only as fall-back code, while at Slang-to-C compilation time the VM uses code written in Slang to generate the C code. One of the side-effect is that the plugin's code is now part of the VM code base and not the client code bases, which is convenient for maintenance since the Cog VM has multiple client languages. However, the implementor now needs to maintain and evolve two version of the primitive, the fall-back code in Smalltalk and the code in Slang. \sd{do we need the next sentence?}This primitive follows the old specification, the full code is available in Appendix~\ref{PluginAppendix}.

@@

\paragraph{Slang+JIT.} \sk{Id like to keep slang+JIT and not slang+RLT as JIT may ring a bell for the non-smalltalkers (while RLT may not). but as we are already using a lot of ST words (primitives, Slang...), does that actually make any sense ?}

@@

 In this section we compare the five different implementations according to the following criteria:
 \begin{itemize}
 	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes). \sd{why not ST, Plugin, VM?} \sk{Yes and No answers to the "recompilation needed" question, so maybe we should keep it that way}
	\item \emph{Implementation language familiar to developers:} Smalltalk application developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code. Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.  We use All if all communities are familiar with the language, VM if all VM developers are and VM-JIT if only a subset of VM developers are.
	\item \emph{Readability:} To evaluate the approach readability, we use the number of lines of code even if this is arguable but used regularly in other papers \cite{xxx}. A better metric would be to use the cyclomatic complexity, however since the  Cog' s RTL generates branches at runtime, it is difficult to use it.
	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to be executed if the primitive fails. This code is maintained by the client language implementors, and not by the VM implementors. Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation\sd{why?}. However, as a VM implementor, we need to maintain what generates the C code of the VM, hence, in the case of Smart Syntax, we need to maintain a different method in each client language. We also count all the Slang implementations as well as the ones on Cog's RTL.
	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
	\item \emph{Optimization potential on large string comparison:}
	C cross-file inlining is not performed by default for internal plugins. External plugins don't have access to the whole VM API and have to go through an interface to do so, leading to an overhead.
	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code. The generated code can therefore differ according the compilers and the optimizations they performed. On the contrary, the JIT will always generate the same native code.
\end{itemize}
